Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews

### 2025-05-29 README.Rmd
resubmit to JAS

1. Install [Docker](https://www.docker.com/get-started/) on your computer 
2. Download the compendium, e.g. from GitHub by clicking on the green `< > Code` button above, then click `Download ZIP`, then unzip on your computer
3. Set your terminal working directory to the compendium, and run `docker build -t wos .` to build the container on your computer, it will take a few minutes and require an internet connection.
4. Run in your terminal `docker run --rm -it -e ROOT=TRUE -e PASSWORD=rstudio -dp 8787:8787 wos` to start the container on your computer
5. Go to http://localhost:8787/ with your browser and enter `rstudio` in both fields to start RStudio in your browser, running in the Docker container with all the required depedencies installed
6. In RStudio, run `rstudioapi::openProject("/home/rstudio/web-of-science-archaeology")` to open the compendium's project. This project uses [`renv`](https://rstudio.github.io/renv/) to document and manage dependencies, this will be activated when the project is first opened in RStudio. 
7. Run ` rmarkdown::render("analysis/paper/paper.qmd", output_dir='/tmp')` to render the Quarto document that is the manuscript submitted for publication
8. After the code has executed, view the rendered output document by running `browseURL("/tmp/paper.html")`
8. When finished, to delete the Docker container from your computer, run this in the terminal `docker ps -aq | xargs docker stop | xargs docker rm`

>>>>

1. Install [Docker](https://www.docker.com/get-started/) on your computer 
2. Download the compendium, e.g. from GitHub by clicking on the green `< > Code` button above, then click `Download ZIP`, then unzip on your computer
3. Set your terminal working directory to the compendium, and run `docker build -t wos .` to build the container on your computer, it will take a few minutes and require an internet connection.
4. Run in your terminal `docker run --rm -it -e ROOT=TRUE -e PASSWORD=rstudio -dp 8787:8787 wos` to start the container on your computer
5. Go to http://localhost:8787/ with your browser and enter `rstudio` in both fields to start RStudio in your browser, running in the Docker container with all the required depedencies installed
6. In RStudio, run `rstudioapi::openProject("/home/rstudio/web-of-science-archaeology")` to open the compendium's project. This project uses [`renv`](https://rstudio.github.io/renv/) to document and manage dependencies, this will be activated when the project is first opened in RStudio. 
7. In RStudio, run `rmarkdown::render("analysis/paper/paper.qmd", output_dir='/tmp')` to render the Quarto document that is the manuscript submitted for publication
8. After the code has executed, view the rendered output document by running in RStudio, `browseURL("/tmp/paper.html")`
9. When finished, to delete the Docker container from your computer, run this in the terminal `docker ps -aq | xargs docker stop | xargs docker rm`

### 2025-05-29 analysis/paper/paper.qmd
resubmit to JAS

---
title: "Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews"
author:
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0003-1689-0557
      institute: uw
institute:
  - uw:
      name: Department of Anthropology, University of Washington
      address: 4218 Memorial Way Northeast, Seattle, WA 98105
fig-dpi: 600
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - templates/scholarly-metadata.lua
  - templates/author-info-blocks.lua
  - templates/pagebreak.lua
bibliography: references.bib
csl: "templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The status of archaeology as a science has been debated for decades and influences how we practice and teach archaeology. This study presents a novel bibliometric assessment of archaeology’s status relative to other fields using a hard/soft framework. It also presents a systematic review of computational reproducibility in published archaeological research. Reproducibility is a factor in the hardness/softness of a field because of its importance in establishing consensus. Analyzing nearly 10,000 articles, I identify trends in authorship, citation practices, and related metrics that position archaeology between the natural and social sciences. A review of reproducibility assessments for the Journal of Archaeological Science reveals persistent challenges, including missing data, unspecified dependencies, and inadequate documentation. To address these issues, I recommend to authors basic practical steps such as standardized project organization and explicit dependency documentation. Strengthening reproducibility will enhance archaeology’s scientific rigor and ensure the verifiability of research findings. This study underscores the urgent need for cultural and technical shifts to establish reproducibility as a cornerstone of rigorous, accountable, and impactful archaeological science.
keywords: |
  Computational reproducibility; Archaeological science; Bibliometric analysis; Open science
highlights: |
  Archaeology occupies an intermediate position between the natural and social sciences.
  Computational reproducibility in archaeology is low, with key barriers in incomplete compendiums, unspecified dependencies, and errors in code.
  Practical steps like logical folder structures, README files, and dependency management tools can substantially improve computational reproducibility.
  A cultural shift toward valuing reproducibility in hiring, promotion, and training is essential for advancing archaeological science.
---

@@

```{r}
#| label: fig-change-over-time
#| fig-cap: Distribution of article characteristics for archaeology articles over time. Data points represent individual articles. The colour of the points indicates if the overall trend is toward softer (orange) or harder (green). Bayesian Generalized Additive Models were computed to fit the lines summarising the relationships between the variables and the time series. For recency of references a Zero-One Inflated Beta distribution family was used with a logit link function. For diversity of references and relative title length (ln) the Gaussian family was used with the canonical identity link function. For the number of authors pages the Negative Binomial family was used with a standard log link function. Further details about the model specifications and diagnostics are presented in the supplementary materials. 
#| fig-height: 4

@@

@fig-aer-summary shows a summary of basic characteristics of the 25 articles that have been through the reproducibility review process so far. The most commonly used software is R, followed by Python. Results generated with proprietary or closed-source software are out of scope for reproducibility reviews. Several distinct types of analyses are well-represented in this sample, especially geometric morphometry, network statistics, and analyses using artificial intelligence or machine learning algorithms (this includes deep learning and neural networks). Most authors are sharing their code and data files via Zenodo, a non-profit generic research data repository hosted by CERN that accepts any file format and freely assigns all publicly available uploads a DOI to make the files easily and uniquely citable [@petersZenodoSpotlightTraditional2017]. In this same category of DOI-issuing, research-grade repositories is OSF (the Open Science Foundation), Figshare, and university repositories. GitHub, a commercial service owned by Microsoft is a code hosting platform that is convenient for collaboration, is also popular among JAS authors, but is a problematic choice because does not offer DOIs or the same commitments to long-term availability as Zenodo. Some authors attached their code and data as journal article supplementary files, but this is a poor choice for long-term availability because these files are typically renamed and converted to different formats during the article production process, making it difficult or impossible for a reader to combine the code and data to reproduce the results.

@@

Documentation is also a key tool is tackling the problems with dependencies described in the previous section. Our finding that dependencies are a common cause of irreproducible results is consistent with previous studies that have identified this as a widespread weakness in communicating computationally intensive research [@trisovicLargescaleStudyResearch2022; @samuelComputationalReproducibilityJupyter2024]. In our sample, issues relating to dependencies are strongly associated with the use of Python. One possible reason for this is that relative to R, Python uses more package managers, more environments, and deeper dependency chains with more complex inter-dependencies that change more rapidly [@decan2016topology; @decanEmpiricalComparisonDependency2019; @korkmazModelingImpactPython2020]. Another reason may be that there is a bigger and more established community of R users in archaeology [@schmidtToolDrivenRevolutionsArchaeological2020; @batistOpenArchaeologyOpen2024] that highly values code that is easy for others to reuse and has evolved practices to effectively communicate dependencies [e.g. @bilottiPointPatternAnalysis2024; @willExploringUtilityUnretouched2025].

The simplest way for archaeologists, to improve here is to write the names and version numbers of the software and packages they used in their README file, as we see in @herskindComputationalLinguisticMethodology2024 and @monnaStudyingSerialityMaterial2024. For more complex research projects, i.e. those using five or more packages or machine learning algorithms, authors should use dependency management tools to keep track of the packages and version numbers needed to reproduce their results. This is an active area of development, and while there are many tools currently available, the most robust and widely used include renv for R [@renv, see examples in @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024] and conda and poetry for Python [@conda2023; @poetry2023].

A more comprehensive solution, and the leading best practice for managing dependencies in many computationally intensive fields using Python in particular, is to include a Dockerfile in the compendium [@moreauContainersComputationalReproducibility2023]. This is a set of a machine- and human-readable instructions that enables a user to recreate the author's computational environment (including those requirements beyond the R or Python packages) on another computer [@nustTenSimpleRules2020]. Dockerfiles are gradually being adopted by archaeologists, see @cremaModellingDiffusionInnovation2024 and @liaoApplyingMaskRCNN2024 for examples, and most of our reproducibility reviews include a recommendation that the author include a Dockerfile to manage complex dependencies efficiently.

@@

A similar debate has been unfolding more broadly about the humanities where Peels and Bouter [-@peelsHumanitiesNeedReplication2018; -@peelsPossibilityDesirabilityReplication2018] have argued that humanities disciplines that use empirical methods should be assessed by how well reanalysis of the original or new data using original or new methods produces the original or equivalent results. Resisting this proposal, @derijckeResistCallsReplicability2018 argue that humanities research is unique because it pursues value and meaning, and a given study can produce multiple valid answers relating to the value and meaning of a study object, so replication is irrelevant as a mark of quality. @peelsEpistemicValuesHumanities2018 disputes this uniqueness, claiming that the humanities has the same epistemic values as the sciences, however some values have more weight in the humanities while others have more weight in the sciences, and unlike in the sciences, humanities scholars often study these epistemic values themselves. A consensus seems to be emerging that for some but not all, studies in the humanities, replication is both possible and desirable, and that replication studies will differ from field to field and might even differ among various studies within a specific field [@bouterHumanitiesNeedReplication2019; @holbrook2019humanities]. A key distinction here is that the locus of evaluation is empirical rather than computational, that is, concerned with appropriate reporting standards and documentation associated with physical evidence (e.g. artefacts and archives) [@stoddenReproducingStatisticalResults2015]. A second important difference is that the debate about qualitative and humanities research is oriented towards *replication* (new data and/or new methods in an independent study to produce the same findings as original publication) rather than *reproducibility* (same data and same methods, e.g. computer code, to produce the same results as original publication) [cf @barbaTerminologiesReproducibleResearch2018]. This orientation to replication emphasizes triangulation techniques that compare and integrate results coming from different traditions, locations, sources and methods, which in turn supports testing whether any given inference is robust in the face of different lines of evidence [@Leonelli2018]. In sum, there are many types of qualitative and humanistic archaeology where it is possible and meaningful to maximize the chances of replication, e.g. by carefully documenting data generating processes, to produce higher quality and more impactful results.

@@

Computational reproducibility is not a panacea; it should not be used as a universally accepted criterion for research quality [@Leonelli2018]. Results that are fully reproducible can contain errors and fraud. It is no guarantee of code quality, or that statistics have been used appropriately [cf. @Vaiglova_2025; Crema 2025], or that data management is consistent with FAIR and CARE principles [@carrollOperationalizingCAREFAIR2021]. It is also time-consuming for authors to ensure their computational work can be reproduced, and for reviewers to evaluate. In a professional environment where job security and career progression is often associated with pressure to publish many high-impact papers, demands for authors to spend time on reproducibility, resulting in less time for publishing more papers, may seem frustrating [@edwards2017academic; @hagstrom1965scientific]. This may seem especially unfair to early career researchers on short-term contracts, who may feel the goalposts for career success are being moved and that they are being asked to do more work their graduate training has not prepared them for. This highlights the need for a culture shift among senior archaeological scientists to value reproducibility in hiring and promotion decision-making. This is important for updating the alignment of quantitative archaeology with normative ideals of scientific practice, such as communal sharing and organized skepticism [@merton1973sociology]. Professors must contribute to this shift by nurturing a culture of reanalysis and reproducibility in their teaching, for example by using replication assignments and by training students in the most current best practices and tools for reproducible research, such as R and Python [@marwickHowUseReplication2020; @dogucuReproducibilityClassroom2025]. A key challenge for the future is changing the dominant habitus (e.g. dispositions, skills, and ways of perceiving) of senior scholars in gatekeeping positions so that reproducibility work will be recognized and rewarded with the same level of symbolic capital afforded to novel high-impact, highly cited publications [@bourdieu1988homo].

>>>>

---
title: "Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews"
author:
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0003-1689-0557
      institute: uw
institute:
  - uw:
      name: Department of Anthropology, University of Washington
      address: 4218 Memorial Way Northeast, Seattle, WA 98105
fig-dpi: 600
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - templates/scholarly-metadata.lua
  - templates/author-info-blocks.lua
  - templates/pagebreak.lua
bibliography: references.bib
csl: "templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The status of archaeology as a science has been debated for decades and influences how we practice and teach archaeology. This study presents a novel bibliometric assessment of archaeology’s status relative to other fields using a hard/soft framework. It also presents a systematic review of computational reproducibility in published archaeological research. Reproducibility is a factor in the hardness/softness of a field because of its importance in establishing consensus. Analyzing nearly 10,000 articles, I identify trends in authorship, citation practices, and related metrics that position archaeology between the natural and social sciences. A survey of reproducibility reviews for the Journal of Archaeological Science reveals persistent challenges, including missing data, unspecified dependencies, and inadequate documentation. To address these issues, I recommend to authors basic practical steps such as standardized project organization and explicit dependency documentation. Strengthening reproducibility will enhance archaeology’s scientific rigor and ensure the verifiability of research findings. This study underscores the urgent need for cultural and technical shifts to establish reproducibility as a cornerstone of rigorous, accountable, and impactful archaeological science.
keywords: |
  Computational reproducibility; Archaeological science; Bibliometric analysis; Open science
highlights: |
  Archaeology occupies an intermediate position between the natural and social sciences.
  Computational reproducibility in archaeology is low, with key barriers in incomplete compendiums, unspecified dependencies, and errors in code.
  Practical steps like logical folder structures, README files, and dependency management tools can substantially improve computational reproducibility.
  A cultural shift toward valuing reproducibility in hiring, promotion, and training is essential for advancing archaeological science.
---

@@

```{r}
#| label: fig-change-over-time
#| fig-cap: Distribution of article characteristics for archaeology articles over time. Data points represent individual articles. The colour of the points indicates if the overall trend is toward softer (orange) or harder (green). Bayesian Generalized Additive Models were computed to fit the lines summarising the relationships between the variables and the time series. For recency of references a Zero-One Inflated Beta distribution family was used with a logit link function. For diversity of references and relative title length (ln) the Gaussian family was used with the canonical identity link function. For the number of authors pages the Negative Binomial family was used with a standard log link function. Further details about the model specifications and diagnostics are presented in the supplementary materials. 
#| fig-height: 6

@@

@fig-aer-summary shows a summary of basic characteristics of the 25 articles that have been through the reproducibility review process so far. The most commonly used software is R, followed by Python. Results generated with proprietary or closed-source software are out of scope for reproducibility reviews. Several distinct types of analyses are well-represented in this sample, especially geometric morphometry, network statistics, and analyses using artificial intelligence or machine learning algorithms (this includes deep learning and neural networks). Most authors are sharing their code and data files via Zenodo, a non-profit generic research data repository hosted by CERN that accepts any file format and freely assigns all publicly available uploads a DOI to make the files easily and uniquely citable [@petersZenodoSpotlightTraditional2017]. In this same category of DOI-issuing, research-grade repositories is OSF (the Open Science Foundation), Figshare, and university repositories. GitHub, a commercial service owned by Microsoft, is a code hosting platform that is convenient for collaboration, is also popular among JAS authors, but is a problematic choice because does not offer DOIs or the same commitments to long-term availability as Zenodo. Some authors attached their code and data as journal article supplementary files, but this is a poor choice for long-term availability because these files are typically renamed and converted to different formats during the article production process, making it difficult or impossible for a reader to combine the code and data to reproduce the results.

@@

Documentation is also a key tool in tackling the problems with dependencies described in the previous section. Our finding that dependencies are a common cause of irreproducible results is consistent with previous studies that have identified this as a widespread weakness in communicating computationally intensive research [@trisovicLargescaleStudyResearch2022; @samuelComputationalReproducibilityJupyter2024]. In our sample, issues relating to dependencies are strongly associated with the use of Python. One possible reason for this is that relative to R, Python uses more package managers, more environments, and deeper dependency chains with more complex inter-dependencies that change more rapidly [@decan2016topology; @decanEmpiricalComparisonDependency2019; @korkmazModelingImpactPython2020]. Another reason may be that there is a bigger and more established community of R users in archaeology [@schmidtToolDrivenRevolutionsArchaeological2020; @batistOpenArchaeologyOpen2024] that highly values code that is easy for others to reuse and has evolved practices to effectively communicate dependencies [e.g. @bilottiPointPatternAnalysis2024; @willExploringUtilityUnretouched2025].

The simplest way for archaeologists to improve here is to write the names and version numbers of the software and packages they used in their README file, as we see in @herskindComputationalLinguisticMethodology2024 and @monnaStudyingSerialityMaterial2024. For more complex research projects, i.e. those using five or more packages or machine learning algorithms, authors should use dependency management tools to keep track of the packages and version numbers needed to reproduce their results. This is an active area of development, and while there are many tools currently available, the most robust and widely used include renv for R [@renv, see examples in @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024] and conda and poetry for Python [@conda2023; @poetry2023].

A more comprehensive solution, and the leading best practice for managing dependencies in many computationally intensive fields using Python in particular, is to include a Dockerfile in the compendium [@moreauContainersComputationalReproducibility2023]. This is a set of machine- and human-readable instructions that enables a user to recreate the author's computational environment (including those requirements beyond the R or Python packages) on another computer [@nustTenSimpleRules2020]. Dockerfiles are gradually being adopted by archaeologists, see @cremaModellingDiffusionInnovation2024 and @liaoApplyingMaskRCNN2024 for examples, and most of our reproducibility reviews include a recommendation that the author include a Dockerfile to manage complex dependencies efficiently.

@@

A similar debate has been unfolding more broadly about the humanities where Peels and Bouter [-@peelsHumanitiesNeedReplication2018; -@peelsPossibilityDesirabilityReplication2018] have argued that humanities disciplines that use empirical methods should be assessed by how well reanalysis of the original or new data using original or new methods produces the original or equivalent results. Resisting this proposal, @derijckeResistCallsReplicability2018 argue that humanities research is unique because it pursues value and meaning, and a given study can produce multiple valid answers relating to the value and meaning of a study object, so replication is irrelevant as a mark of quality. @peelsEpistemicValuesHumanities2018 disputes this uniqueness, claiming that the humanities has the same epistemic values as the sciences, however some values have more weight in the humanities while others have more weight in the sciences, and unlike in the sciences, humanities scholars often study these epistemic values themselves. A consensus seems to be emerging that for some but not all, studies in the humanities, replication is both possible and desirable, and that replication studies will differ from field to field and might even differ among various studies within a specific field [@bouterHumanitiesNeedReplication2019; @holbrook2019humanities].

A key distinction here is that the locus of evaluation is empirical rather than computational, that is, concerned with appropriate reporting standards and documentation associated with physical evidence (e.g. artefacts and archives) [@stoddenReproducingStatisticalResults2015]. A second important difference is that the debate about qualitative and humanities research is oriented towards *replication* (new data and/or new methods in an independent study to produce the same findings as original publication) rather than *reproducibility* (same data and same methods, e.g. computer code, to produce the same results as original publication) [cf @barbaTerminologiesReproducibleResearch2018]. This orientation to replication emphasizes triangulation techniques that compare and integrate results coming from different traditions, locations, sources and methods, which in turn supports testing whether any given inference is robust in the face of different lines of evidence [@Leonelli2018]. In sum, there are many types of qualitative and humanistic archaeology where it is possible and meaningful to maximize the chances of non-computational replication, e.g. by carefully documenting data generating processes, to produce higher quality and more impactful results.

@@

Computational reproducibility is not a panacea; it should not be used as a universally accepted criterion for research quality [@Leonelli2018]. Results that are fully reproducible can contain errors and fraud. It is no guarantee of code quality, or that statistics have been used appropriately [cf. @Vaiglova_2025; @crema_2025], or that data management is consistent with FAIR and CARE principles [@carrollOperationalizingCAREFAIR2021]. It is also time-consuming for authors to ensure their computational work can be reproduced, and for reviewers to evaluate. This is especially the case for papers reporting results generated by long-running simulation or deep learning. These may be impracticable to fully reproduce by a typical peer reviewer who does not have access to specialised computing facilities. In a professional environment where job security and career progression is often associated with pressure to publish many high-impact papers, demands for authors to spend time on reproducibility, resulting in less time for publishing more papers, may seem frustrating [@edwards2017academic; @hagstrom1965scientific]. This may seem especially unfair to early career researchers on short-term contracts, who may feel the goalposts for career success are being moved and that they are being asked to do more work their graduate training has not prepared them for. This highlights the need for a culture shift among senior archaeological scientists to value reproducibility in hiring and promotion decision-making. This is important for updating the alignment of quantitative archaeology with normative ideals of scientific practice, such as communal sharing and organized skepticism [@merton1973sociology]. Professors must contribute to this shift by nurturing a culture of reanalysis and reproducibility in their teaching, for example by using replication assignments and by training students in the most current best practices and tools for reproducible research, such as R and Python [@marwickHowUseReplication2020; @dogucuReproducibilityClassroom2025]. A key challenge for the future is changing the dominant habitus (e.g. dispositions, skills, and ways of perceiving) of senior scholars in gatekeeping positions so that reproducibility work will be recognized and rewarded with the same level of symbolic capital afforded to novel high-impact, highly cited publications [@bourdieu1988homo].
